Simulations of gravity-driven flow of binary liquids in microchannels

### 2011-03-25 Text/paper.tex
Adding details for revisions.

\section{Results}
\subsection{The nondimensionalization and initialization procedure}
We now introduce the procedure of nondimensionalization, as well as
the initialization technique we used to determine the parameters necessary for
the simulation. The capillary number defined in Eq.~\ref{capillary:number:definition} governs the
interface thickness.
%The bubble velocity $U_{\mathrm{bubble}}$ is taken in the center of the bubble where the film
%thickness is examined.
While parameters such as $\mu_{\mathrm{liq}}$, $\gamma$ and the capillary number $\Ca$ are given
or defined through the model parameters, the bubble velocity $U_{\mathrm{bubble}}$ is not given
explicitly.
In our simulations we approximate $U_{bubble}$ with the
Poiseuille profile center velocity, which is given as:
\begin{equation}
\begin{aligned}
U_{\mathrm{bubble}} \approx U_{\mathrm{liq}}=\frac{H_{\mathrm{eff}}^2}{8
\mu_{\mathrm{liq}}}\frac{\mathrm{d}P}{\mathrm{d}x}\\
U_{\mathrm{bubble}}\approx
\frac{{(N_y-2)}^2}{\frac{8}{6}(\tau_{\mathrm{liq}}-\frac{1}{2})}\frac{\mathrm{d}P}{\mathrm{d}
x } ,
\end{aligned}
\end{equation}
where $N_y$ is the grid number in the vertical direction. As will be discussed further, the
effective channel width is defined as $H_{\mathrm{eff}}=N_y-2$ because of bounce-back nodes to
mimic no-slip conditions at the walls.
For the Poiseuille profile one can estimate the body force (pressure gradient) through the
capillary number and grid resolution:
\begin{equation}
\label{poiseuille:velocity:center}
\begin{aligned}
&\frac{\mathrm{d}P}{\mathrm{d}x}=\frac{8}{{(N_y-2)}^2}\gamma \Ca.
\end{aligned}
\end{equation}
In simulations the body force (\ref{poiseuille:velocity:center}) causes the bubble
velocity to differ from that implied by the Poiseuille relation because of the bubble presence in a
flow. That is
why an iteration procedure is needed to achieve the 
desired velocity and 
capillary number. 
However, the initialization body force given by (\ref{poiseuille:velocity:center}) is a good
starting approximation.

@@

Overall, the setting simulations procedure can be described in steps as follows:
\begin{description}
 \item[Capillary number] One first needs to be supplied with the capillary number for
simulations. For the demonstration purposes we chose $\Ca=0.05$. 
 \item[Film thickness] After the capillary number is prescribed, one needs to approximate the film
thickness $\delta$. It can be done either by taking correlations from numerical simulations by
\citet{heil-bretherton} and \citet{giavedoni-numerical} or by using the classical Bretherton
correlation for small cappillary numbers $\delta=0.6687 \Ca^{2/3}$. In the case of $\Ca=0.05$, the 
film thickness is taken from {\color{green} Fig.~\ref{fig:heil:planar}},
  and for Reynolds number $40$ equals $0.06$.  
\item[Grid choice] After pecifying the film thickness, one needs to choose the associate
number of nodes to resolve the film thickness. Note that for paremeters indicated above the
interface thickness is spread over approximately $5$ lattice units. The study of the interface
resolution to the film thickness, Section \ref{section:grid:refinement}, suggests to choose the
film thickness to be $2-2.5$ times larger than the interface thickness. Therefore, we want to
  resolve $6\%$ of the whole channel width with $12$ lattice units.  Thus, the channel
  width is $200$ lattice units. We take the bubble length equal to $5$ channel widths for the film
thickness to establish
  and the distance between bubbles as $3$ bubble lengths to avoid mutual influence of bubbles on
each other. Thus, the whole grid size is $200 \times 3000$.
 \item[Velocity] The relaxation parameters are taken to be in the stable range for simulations.
In the present simulations the viscosity ratio is $\dfrac{\mu_{liq}}{\mu_{gas}}=10$ with the
correspoding relaxation times $\tau_{liq}=2.5$ and $\tau_{gas}=0.7$. Given the viscosity of the
liquid one can obtain the velocity of the bubble from the capillary number: 
\begin{equation}
\begin{aligned}
&U_{\mathrm{bubble}}=\Ca \frac{\gamma}{\mu_{\mathrm{liq}}}=\Ca \frac{\sqrt{\frac{8 k
A}{9}}}{\frac{1}{3}(\tau_{\mathrm{liq}}-\frac{1}{2})}\\
&U_{\mathrm{bubble}}=0.05 \frac{0.0377}{0.6666}=2.82 \cdot10^{-3},
\end{aligned}
\end{equation}
where $\mu_{\mathrm{liq}}=1/3 (2.5-0.5)=0.6666$.
\item[Body force] 
  The velocity implies the following pressure gradient calculated from the
  Poiseuille flow solution:
  \begin{equation}
  \begin{aligned}
  &\frac{\mathrm{d}P}{\mathrm{d}x}=\frac{8}{N_y^2}\gamma \Ca=3.77\cdot10^{-7}.
  \end{aligned}
  \end{equation}

After resolving the interface one can  calculate the
  velocity from the capillary number as:
  This velocity implies the following pressure gradient calculated from the
  Poiseuille flow solution, Eq. \ref{poiseuille:velocity:center}: 
  \begin{equation}
  \begin{aligned}
  \frac{\mathrm{d}P}{\mathrm{d}x}=9.42 \cdot 10^{-9}.
  \end{aligned}
  \end{equation}
Note that the velocity is proportional to the physical time step (small velocities imply longer
running
simulations):
\begin{equation}
\Delta t =\frac{U_{\mathrm{LBM}}}{U_{\mathrm{phys}}} \Delta x ,
\end{equation}
where, for example, the real velocities in the physical world can reach
approximately $1$ m/s and the real height up to $1$ mm. In this case one
can easily approximate the physical time step during one lattice Boltzmann iteration:
\begin{equation}
\Delta t = 7.05 \cdot 10^{-10} \,\mathrm{s}
\end{equation}
which is small. A typical simulation runs approximately
$100000-300000$ time steps due to reach steady-state. That gives 
physical time simulation around $10^{-5}-10^{-4}\,\mathrm{s}$, which is
probably not enough for the physical system to come to the pseudo-steady state.

@@

This gives a
grid size of $400 \times 6000$. In terms of computer memory
requirements, the calculations with double precision accuracy demand:
\begin{equation}
\begin{aligned}
&2\text{ sets of populations } \times \,\, 8\,\, \text{bytes per population
} \times \,\, 9 \text{
populations}\\
&\times\,400\,\times\,6000\approx 345 \mathrm{MB}. 
\end{aligned}
\end{equation}
Even with an interface thickness of $62.5$ percent  of the
film thickness, one gets (in 2D) a memory
requirement of $345 \mathrm{MB}$.

   The parameters in the previous subsection imply that long
computations are required in order for the system to come to equilibrium. The simulation becomes
longer with a decrease of the
capillary number. Bretherton obtained an analytical solution for the range of low capillary
numbers $Ca \leq 0.005$. However, many researchers extended it to the range of capillary numbers
greater than
$0.005$, and to higher Reynolds numbers. One can refer to figures
  \ref{fig:giavedoni:planar} and \ref{fig:heil:planar} taken from the work of
  \citet{giavedoni-numerical} and \citet{heil-bretherton}, who cross validated their
  simulations against the Bretherton data and each other. Even though the LBM can simulate
flows with small capillary numbers, i.e. less than $0.005$, the simulation time
is relatively long. Therefore, in our work we focus on the range of capillary numbers $\Ca$
employed by the simulations of
\citet{giavedoni-numerical} who cross validated
results for all ranges of capillary numbers.  Our intention is
  to show physically correct behavior for moderate $Ca$,
which
  can be easily and quickly validated on a computer.
  One can see from Fig.~\ref{fig:heil:planar} that the film thickness for the
  capillary number $0.05$ ranges from $0.12$ to $0.15$ in a wide Reynolds
  number range (from $0$ to $300$).

@@

\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/giavedoni_planar.eps}
\caption{\citet{giavedoni-numerical} gathered results across the
literature for different capillary numbers (reprinted from \cite{giavedoni-numerical}). Results are
scaled to the half-width of the
channel. $h_{\infty}$ is the film thickness of air finger at
the infinite distance from the front meniscus. \label{fig:giavedoni:planar}}
\includegraphics[width=0.97\textwidth]{Figures/heil-planar.eps}
\caption{\citet{heil-bretherton} performed simulations for different ranges of
capillary numbers and Reynolds numbers (reprinted from \cite{heil-bretherton}). Results
are scaled to the half-width of the channel. $h_{0}$ is the film thickness of air finger at
the infinite distance from the front meniscus.
\label{fig:heil:planar}}
\end{figure}
Note that we base the initialization techniques on correlations for the
capillary number and the body force. However, in the binary liquid framework
those correlations are approximations. The strongest assumption is
in calculating the body force from the Poiseuille velocity profile.
In reality, the less viscous bubble moves faster than the surrounding liquid. By
substituting the obtained velocity in the capillary number equation, a
larger capillary number can be obtained. In practice, one needs to take the bubble velocity from the
simulations and
recalculate all the necessary flow characteristics, i.e. adjust the
 applied force to obtain the desired velocity/capillary number.

\subsection{Grid refinement}
\label{section:grid:refinement}
To properly estimate the interface resolution one needs to study the convergence as a function of
the grid resolution. To do that the grid resolution is varied while all remaining parameters,
including the bubble velocity and the capillary number, are fixed.  Our goal is to determine the 
ratio of the interface thickness to the 
film thickness at which results are no longer dependent on the grid resolution.

@@

After choosing the reference parameters, the grid refinement procedure
needs to keep the macroscopic parameters constant.  It is easy to check
that it yields the following quantity to
be independent of grid size
$H_{\mathrm{eff}}^2\frac{\mathrm{d}P}{\mathrm{d}x}=(N_y-2)^2\frac{\mathrm{d}P}{\mathrm{d}
x } = \mathrm{const}$. We performed a number of simulations for the prescribed grids. The
simulation results in terms of grid numbers $N_x$, $N_y$, the film thickness $\delta$, center bubble
velocity $U_{bubble}$ are summarized in Table \ref{table:parameters:grid:refinement}.
The unified scaled profiles are shown in Fig. \ref{fig:grid:profiles}. One can
see that results converge for $H_{\mathrm{eff}}\geq 175$, and that with proper
initialization techniques, large enough time and different wall
wettabilities, results are different only in the 3rd digit even for
underresolved film thicknesses. To calculate how well the interface is
resolved, the ratio of the interface thickness to the film thickness is calculated. The
interface itself occupies approximately $5 \xi$, where
$\xi=\sqrt{k/A}=1$. The ratio of the interface thickness to the film thickness
$5\xi/H_{\mathrm{film}}$
 is shown in Table \ref{table:parameters:grid:refinement}. Based on these results one can conclude that
 the interface needs to be resolved as $40-50$ percent of the
expected film thickness for simulations to be grid independent. We further examine the
velocities in the center of the bubble to calculate the capillary number. One can see from Table
\ref{table:parameters:grid:refinement} that the bubble velocities are consistent and the
calculated capillary number corresponding to these velocities is $0.047$, which is {\color{red}
close to} the capillary number we aimed for. The corresponding difference can be
attributed to the Poiseuille profile body force initialization.% and the
 %density and viscosity ratios, which are much larger in the
%Bretherton problem.
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/Grid/norm_grid_profs.eps}
\caption{Grid-refined profiles for the effective
channel widths
$H_{\mathrm{eff}}=100,125,150,175,200,225$.$\delta$ is scaled on $H_{\mathrm{eff}}$ and $\delta=0$
corresponds to the wall location. The profiles were taken at $x=14$ (nondimensional coordinates).
Capillary number is $\Ca=0.05$. The capillary number obtained from simulations $\Ca=0.047$. 
%One can seeconvergent results even for the underresolved film
%thickness.
\label{fig:grid:profiles}}
\end{figure}
%Here will be a consolidated table
\begin{table}
\begin{tabularx}{0.97\textwidth}{|X|X|X|X|X|X|X|}
\hline
$N_x$&$N_y$&$\delta$&$U_\mathrm{bubble}$&$\frac{5\xi}{H_{\mathrm{film}}}$&$N_\mathrm{iter}$&{
\color{red} $Re$}\\
\hline
$1500$&$102$&$0.0694$&$0.0028$&$0.824$&$200000$&$0.44$\\
\hline
$1875$&$127$&$0.0690$&$0.0028$&$0.646$&$250000$&$0.53$\\
\hline
$2250$&$152$&$0.0675$&$0.0027$&$0.539$&$300000$&$0.63$\\
\hline
$2625$&$177$&$0.0674$&$0.0027$&$0.453$&$350000$&$0.73$\\
\hline
$3000$&$202$&$0.0662$&$0.0027$&$0.400$&$400000$&$0.83$\\
\hline
$3375$&$227$&$0.0664$&$0.0026$&$0.355$&$450000$&$0.91$\\
\hline
\end{tabularx}
\caption{The parameters and results for grid resolution. The simulated domain is
of size $N_x \times N_y$. $U_\mathrm{bubble}$ is the velocity of the interface at the fron tip of
the bubble.  $5\xi$ is the interface thickness. $H_{\mathrm{film}}=\delta (N_y-2)$ is the size of
the film in
lattice Boltzmann units. $Ca=0.05$ was taken to initialize the simulation.
\label{table:parameters:grid:refinement}}
\end{table}

@@

\subsection{Capillary number region}
The purpose of this section is to validate the correlations of
\citet{giavedoni-numerical} in Fig. \ref{fig:giavedoni:planar} and
\citet{heil-bretherton} in Fig. \ref{fig:heil:planar} for a range of capillary
numbers. Because of limited computational resources, we skip the
small capillary numbers and make calculations for the range of capillary numbers $0.03-0.8$, which
is a computationally reasonable task.  To be consistent with the previous sections, we choose the grid to be
$202 \times 3001$. Then $5$ lattice Boltzmann units do not occupy more than $60$
percent of the effective film thickness. Because the simulation gets unstable with
smaller grids and larger gradients, all the capillary number simulations were
performed on the same grid. To properly initialize the body force, the
proportionality law was utilized. The forcing
$6 \cdot 10^{-6}/16$ was chosen to obtain the predicted capillary
number $0.05$. The forcing for other grids can be obtained using
simple proportionality relationships:
&\Ca_{\mathrm{lit}} \propto U_{\mathrm{bubble}}\\
&U_{\mathrm{bubble}} \propto \frac{\mathrm{d}P}{\mathrm{d}x} N_y^2\\
&\Ca_{\mathrm{lit}} \propto \frac{\mathrm{d}P}{\mathrm{d} x} N_y^2 \text{ or }\\
&\frac{\mathrm{d}P}{\mathrm{d} x} \propto \frac{\Ca_{\mathrm{lit}}}{N_y^2},
where the subscript ,,lit'' stands for the predicted capillary number
\cite{giavedoni-numerical,heil-bretherton}. The grid number $N_y$ is not
involved because all simulations are conducted on the same grid.
The pressure gradient can be obtained through the capillary number
ratio:
\frac{\mathrm{d}P}{\mathrm{d} x}=6 \cdot 10^{-6}/16 \frac{\Ca_{\mathrm{lit}}}{0.05}
The film thickness is initialized through the ratio of capillary numbers as well:
\begin{equation*}
w=12 \frac{\Ca_{\mathrm{lit}}}{0.05}
\end{equation*}
The results obtained after $2\cdot10^5$ steps are presented in Table
\ref{table:parameters:capillary:number}. 
{\color{red} We chose to measure the film thickness in the middle of the bubble. The
classical Bretherton formulation is for the measurement of the film thickness at ``infinity''.
However, \citet{cerro-bubble-train} indicate that for the gravity-driven bubble train flow in tubes
it is enough for bubble to have a length of two-three diameters to apply analysis for very long
bubbles. In comparison with the axisymmetric circular geometry, for three-dimensional square shaped
microchannels \citet{heil-threedim} indicate the measurement of the film thickness at $5.5$
channel heights from the front bubble tip. This distance was shown sufficient for the
film thickness to establish itself for the capillary number range $\Ca<4$.
\citet{giavedoni-numerical} showed that the film stabilizes at the distances as $2.6-4.0$ diameters
from the front tip depending on the Reynolds number.  We chose to measure the film thickness in the
middle of the bubble, which is located at least at the distance of $2.5-3$ channel heights from the
front bubble tip. Moreover, the film thickness examination, see Section
\ref{section:film:thickness:variation}, shows that for the capillary range of interest, i.e.
$0.05\leq \Ca \leq 1$, the standard deviation from the bubble middle film thickness is around $7$
percent for small capillary numbers ($\Ca=0.05$) and less than $1$ percent for larger ones ($\Ca
\geq 0.1$) at the distances ranging from one channel height from the rear tip upto one channel
height to the front tip. This certainly validates the point of choice to measure bubble thickness.}
One can see that the calculated
capillary numbers are overpredicted but the obtained thicknesses are overpredicted
as well due to the body force setting caused by the Poiseuille flow assumption. If one
needs to obtain the desired capillary number, the shooting method for the forcing is
necessary, starting with the Poiseuille pressure gradient force as the initial condition. We extracted
the data of 
\citet{giavedoni-numerical} and \citet{heil-bretherton} and compared them with our results (see 
Fig. \ref{fig:capillary:comparison}).
\includegraphics[width=0.97\textwidth]{Figures/Capillary/capillaries_comparison_real.eps}
\caption{A comparison between simulation results and results of
\citet{giavedoni-numerical} and \citet{heil-bretherton}. One can see a
reasonable agreement. The plots depict the film thickness as a function of the
the capillary number.\label{fig:capillary:comparison}}

>>>>

\section{Results}
\subsection{The nondimensionalization and initialization procedure}
\label{section:init}
We now introduce the procedure of nondimensionalization, as well as
the initialization technique we used to determine the parameters necessary for
the simulation. The capillary number defined in Eq.~\ref{capillary:number:definition} governs the
interface thickness.
%The bubble velocity $U_{\mathrm{bubble}}$ is taken in the center of the bubble where the film
%thickness is examined.
While parameters such as $\mu_{\mathrm{liq}}$, $\gamma$ and the capillary number $\Ca$ are given
or defined through the model parameters, the bubble velocity $U_{\mathrm{bubble}}$ is not given
explicitly.
In our simulations we approximate $U_{bubble}$ with the
Poiseuille profile center velocity, which is given as:
\begin{equation}
\begin{aligned}
U_{\mathrm{bubble}} \approx U_{\mathrm{liq}}=\frac{H_{\mathrm{eff}}^2}{8
\mu_{\mathrm{liq}}}\frac{\mathrm{d}P}{\mathrm{d}x}\\
U_{\mathrm{bubble}}\approx
\frac{{(N_y-2)}^2}{\frac{8}{6}(\tau_{\mathrm{liq}}-\frac{1}{2})}\frac{\mathrm{d}P}{\mathrm{d}
x } ,
\end{aligned}
\end{equation}
where $N_y$ is the grid number in the vertical direction. As will be discussed further, the
effective channel width is defined as $H_{\mathrm{eff}}=N_y-2$ because of bounce-back nodes to
mimic no-slip conditions at the walls.
For the Poiseuille profile one can estimate the body force (pressure gradient) through the
capillary number and grid resolution:
\begin{equation}
\label{poiseuille:velocity:center}
\begin{aligned}
&\frac{\mathrm{d}P}{\mathrm{d}x}=\frac{8}{{(N_y-2)}^2}\gamma \Ca.
\end{aligned}
\end{equation}
In simulations the body force (\ref{poiseuille:velocity:center}) causes the bubble
velocity to differ from that implied by the Poiseuille relation because of the bubble presence in a
flow. That is
why an iteration procedure is needed to achieve the 
desired velocity and 
capillary number. 
However, the initialization body force given by (\ref{poiseuille:velocity:center}) is a good
starting approximation.

@@

Overall, the setting simulations procedure can be described in steps as follows:
\begin{description}
 \item[Capillary number] One first needs to be supplied with the capillary number for
simulations. For the demonstration purposes we chose $\Ca=0.05$. 
 \item[Film thickness] After the capillary number is prescribed, one needs to approximate the film
thickness $\delta$. It can be done either by taking correlations from numerical simulations by
\citet{heil-bretherton} and \citet{giavedoni-numerical} or by using the classical Bretherton
correlation for small cappillary numbers $\delta=0.6687 \Ca^{2/3}$. In the case of $\Ca=0.05$, the 
film thickness is taken from {\color{green} Fig.~\ref{fig:heil:planar}},
  and for Reynolds number $40$ equals $0.06$.  
\item[Grid choice] After pecifying the film thickness, one needs to choose the associate
number of nodes to resolve the film thickness. Note that for paremeters indicated above the
interface thickness is spread over approximately $5$ lattice units. The study of the interface
resolution to the film thickness, Section \ref{section:grid:refinement}, suggests to choose the
film thickness to be $2-2.5$ times larger than the interface thickness. Therefore, we want to
  resolve $6\%$ of the whole channel width with $12$ lattice units.  Thus, the channel
  width is $200$ lattice units. We take the bubble length equal to $5$ channel widths for the film
thickness to establish
  and the distance between bubbles as $3$ bubble lengths to avoid mutual influence of bubbles on
each other. Thus, the whole grid size is $200 \times 3000$.
 \item[Velocity] The relaxation parameters are taken to be in the stable range for simulations.
In the present simulations the viscosity ratio is $\dfrac{\mu_{liq}}{\mu_{gas}}=10$ with the
correspoding relaxation times $\tau_{liq}=2.5$ and $\tau_{gas}=0.7$. Given the viscosity of the
liquid one can obtain the velocity of the bubble from the capillary number: 
&U_{\mathrm{bubble}}=\Ca \frac{\gamma}{\mu_{\mathrm{liq}}}=\Ca \frac{\sqrt{\frac{8 k
A}{9}}}{\frac{1}{3}(\tau_{\mathrm{liq}}-\frac{1}{2})}\\
&U_{\mathrm{bubble}}=0.05 \frac{0.0377}{0.6666}=2.82 \cdot10^{-3},
where $\mu_{\mathrm{liq}}=1/3 (2.5-0.5)=0.6666$.
\item[Body force] 
Thus, it is desired to obtain the prescribed velocity $U_{\mathrm{bubble}}$ in the simulations. The
flow in the simulations is conducted by imposing the body force. Thus, the assumption is needed of
how the body force is connected with the bubble velocity $U_{\mathrm{bubble}}$. We assume that the
flow is close to the Poiseuille flow:
\frac{\mathrm{d}P}{\mathrm{d}x}=\frac{8\mu}{H_{\mathrm{eff}}^2}
U_{\mathrm{bubble}}=\frac{8}{N_y^2}\gamma \Ca=3.77\times 10^{-7}.  
After imposing the body force the simulations can be run. A typical simulation runs approximately
$100000-300000$ time steps due to reach steady-state.
%Note that the velocity is proportional to the physical time step (small velocities imply longer
%running
%simulations):
%\begin{equation}
%\Delta t =\frac{U_{\mathrm{LBM}}}{U_{\mathrm{phys}}} \Delta x ,
%\end{equation}
%where, for example, the real velocities in the physical world can reach
%approximately $1$ m/s and the real height up to $1$ mm. In this case one
%can easily approximate the physical time step during one lattice Boltzmann iteration:
%\begin{equation}
%\Delta t = 7.05 \cdot 10^{-10} \,\mathrm{s}
%\end{equation}
%which is small.  That gives 
%physical time simulation around $10^{-5}-10^{-4}\,\mathrm{s}$, which is
%probably not enough for the physical system to come to the pseudo-steady state.

@@

@@

Note that following the prescribed procedure one can see that a computer memory requirements are
high for small capillary numbers due to large grids to resolve the film thickness. Thus, the
simulations in the present work cover range of capillary numbers from $0.05$ to $1.0$.
Although, Bretherton obtained an analytical solution for the range of low capillary
numbers $Ca \leq 0.005$, many researchers, for example \citet{giavedoni-numerical} and
\citet{heil-bretherton}, extended it to the range of capillary numbers
greater than
$0.005$, and to higher Reynolds numbers. %One can refer to figures
%  \ref{fig:giavedoni:planar} and \ref{fig:heil:planar} taken from the work of
%  \citet{giavedoni-numerical} and \citet{heil-bretherton}, who cross validated their
% simulations against the Bretherton data and each other.
%Even though the LBM can simulate
%flows with small capillary numbers, i.e. less than $0.005$, the simulation time
%is relatively long. 
Therefore, the focus of our work on the capillary range larger $\Ca\geq0.05$ is legitimate as far
as we base it on comparison with already cross validated
results for all ranges of capillary numbers.  Our intention is
  to show physically correct behavior for moderate $Ca$,
which
  can be easily and quickly validated on a computer.
%One can see from Fig.~\ref{fig:heil:planar} that the film thickness for the
%  capillary number $0.05$ ranges from $0.12$ to $0.15$ in a wide Reynolds
% number range (from $0$ to $300$).
%This gives a
%grid size of $400 \times 6000$. In terms of computer memory
%requirements, the calculations with double precision accuracy demand:
%\begin{equation}
%\begin{aligned}
%&2\text{ sets of populations } \times \,\, 8\,\, \text{bytes per population
%} \times \,\, 9 \text{
%populations}\\
%&\times\,400\,\times\,6000\approx 345 \mathrm{MB}. 
%\end{aligned}
%\end{equation}
%Even with an interface thickness of $62.5$ percent  of the
%film thickness, one gets (in 2D) a memory
%requirement of $345 \mathrm{MB}$. 
\includegraphics[width=0.97\textwidth]{Figures/giavedoni_planar.eps}
\caption{\citet{giavedoni-numerical} gathered results across the
literature for different capillary numbers (reprinted from \cite{giavedoni-numerical}). Results are
scaled to the half-width of the
channel. $h_{\infty}$ is the film thickness of air finger at
the infinite distance from the front meniscus. \label{fig:giavedoni:planar}}
\includegraphics[width=0.97\textwidth]{Figures/heil-planar.eps}
\caption{\citet{heil-bretherton} performed simulations for different ranges of
capillary numbers and Reynolds numbers (reprinted from \cite{heil-bretherton}). Results
are scaled to the half-width of the channel. $h_{0}$ is the film thickness of air finger at
the infinite distance from the front meniscus.
\label{fig:heil:planar}}

Note that we base the initialization techniques on correlations for the
capillary number and the body force. However, in the binary liquid framework
those correlations are approximations. The strongest assumption is
in calculating the body force from the Poiseuille velocity profile.
In reality, the less viscous bubble moves faster than the surrounding liquid. By
substituting the obtained velocity in the capillary number equation, a
larger capillary number can be obtained. In practice, one needs to take the bubble velocity from the
simulations and
recalculate all the necessary flow characteristics, i.e. adjust the
 applied force to obtain the desired velocity/capillary number. However, simulations show that the
Poiseuille flow assumptions work reasonably well.

{\color{red}
Note that the initialization procedure is a good guide of designing the simulations. For example
for the grid refinement, see Section \ref{section:grid:refinement}, we keep the same velocity
$U_{\mathrm{bubble}}$ and $\Ca$. This implies the conservation of the quantity
$H_{\mathrm{eff}}^2\frac{\mathrm{d}P}{\mathrm{d}x}=(N_y-2)^2\frac{\mathrm{d}P}{\mathrm{d}
x } = \mathrm{const}$. For the designing capillary range simulations, we use a proportionality law
to initialize body forces knowing results of just one simulation:
\begin{equation}
\begin{aligned}
&\Ca_{\mathrm{lit}} \propto U_{\mathrm{bubble}}\\
&U_{\mathrm{bubble}} \propto \frac{\mathrm{d}P}{\mathrm{d}x} H_{\mathrm{eff}}^2\\
&\Ca_{\mathrm{lit}} \propto \frac{\mathrm{d}P}{\mathrm{d} x} H_{\mathrm{eff}}^2 \text{ or }\\
&\frac{\mathrm{d}P}{\mathrm{d} x} \propto \frac{\Ca_{\mathrm{lit}}}{H_{\mathrm{eff}}^2},
\end{aligned}
\end{equation}
where the subscript ,,lit'' stands for the predicted capillary number
\cite{giavedoni-numerical,heil-bretherton}. The grid number $N_y$ is not
involved because all simulations are conducted on the same grid.
}
\subsection{Grid refinement}
\label{section:grid:refinement}
To properly estimate the interface resolution one needs to study the convergence as a function of
the grid resolution. To do that the grid resolution is varied while all remaining parameters,
including the bubble velocity and the capillary number, are fixed.  Our goal is to determine the 
ratio of the interface thickness to the 
film thickness at which results are no longer dependent on the grid resolution.

@@

{\color{red} After choosing the reference parameters, the grid refinement procedure
needs to keep the macroscopic parameters constant. One can refer to Section \ref{section:init}.}  
%It is easy to check
%that it yields the following quantity to
%be independent of grid size
%$H_{\mathrm{eff}}^2\frac{\mathrm{d}P}{\mathrm{d}x}=(N_y-2)^2\frac{\mathrm{d}P}{\mathrm{d}
%x } = \mathrm{const}$. 
We performed a number of simulations for the prescribed grids. The
simulation results in terms of grid numbers $N_x$, $N_y$, the film thickness $\delta$, center bubble
velocity $U_{bubble}$ are summarized in Table \ref{table:parameters:grid:refinement}.
The unified scaled profiles are shown in Fig. \ref{fig:grid:profiles}. One can
see that results converge for $H_{\mathrm{eff}}\geq 175$, and that with proper
initialization techniques, large enough time and different wall
wettabilities, results are different only in the 3rd digit even for
underresolved film thicknesses. To calculate how well the interface is
resolved, the ratio of the interface thickness to the film thickness is calculated. The
interface itself occupies approximately $5 \xi$, where
$\xi=\sqrt{k/A}=1$. The ratio of the interface thickness to the film thickness
$5\xi/H_{\mathrm{film}}$
 is shown in Table \ref{table:parameters:grid:refinement}. Based on these results one can conclude that
 the interface needs to be resolved as $40-50$ percent of the
expected film thickness for simulations to be grid independent. We further examine the
velocities in the center of the bubble to calculate the capillary number. One can see from Table
\ref{table:parameters:grid:refinement} that the bubble velocities are consistent and the
calculated capillary number corresponding to these velocities is $0.047$, which is {\color{red}
close to} the capillary number we aimed for. The corresponding difference can be
attributed to the Poiseuille profile body force initialization.% and the
 %density and viscosity ratios, which are much larger in the
%Bretherton problem.
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/Grid/norm_grid_profs.eps}
\caption{Grid-refined profiles for the effective
channel widths
$H_{\mathrm{eff}}=100,125,150,175,200,225$.$\delta$ is scaled on $H_{\mathrm{eff}}$ and $\delta=0$
corresponds to the wall location. The profiles were taken at $x=14$ (nondimensional coordinates).
Capillary number is $\Ca=0.05$. The capillary number obtained from simulations $\Ca=0.047$. 
%One can seeconvergent results even for the underresolved film
%thickness.
\label{fig:grid:profiles}}
\end{figure}
%Here will be a consolidated table
\begin{table}
\begin{tabularx}{0.97\textwidth}{|X|X|X|X|X|X|X|}
\hline
$N_x$&$N_y$&$\delta$&$U_\mathrm{bubble}$&$\frac{5\xi}{H_{\mathrm{film}}}$&$N_\mathrm{iter}$&{
\color{red} $Re$}\\
\hline
$1500$&$102$&$0.0694$&$0.0028$&$0.824$&$200000$&$0.44$\\
\hline
$1875$&$127$&$0.0690$&$0.0028$&$0.646$&$250000$&$0.53$\\
\hline
$2250$&$152$&$0.0675$&$0.0027$&$0.539$&$300000$&$0.63$\\
\hline
$2625$&$177$&$0.0674$&$0.0027$&$0.453$&$350000$&$0.73$\\
\hline
$3000$&$202$&$0.0662$&$0.0027$&$0.400$&$400000$&$0.83$\\
\hline
$3375$&$227$&$0.0664$&$0.0026$&$0.355$&$450000$&$0.91$\\
\hline
\end{tabularx}
\caption{The parameters and results for grid resolution. The simulated domain is
of size $N_x \times N_y$. $U_\mathrm{bubble}$ is the velocity of the interface at the fron tip of
the bubble.  $5\xi$ is the interface thickness. $H_{\mathrm{film}}=\delta (N_y-2)$ is the size of
the film in
lattice Boltzmann units. $Ca=0.05$ was taken to initialize the simulation.
\label{table:parameters:grid:refinement}}
\end{table}

@@

\subsection{Capillary number region}
\label{section:capillary:range}
The purpose of this section is to validate the correlations of
\citet{giavedoni-numerical} in Fig. \ref{fig:giavedoni:planar} and
\citet{heil-bretherton} in Fig. \ref{fig:heil:planar} for a range of capillary
numbers. Because of limited computational resources, we skip the
small capillary numbers and make calculations for the range of capillary numbers $0.03-0.8$, which
is a computationally reasonable task.  To be consistent with the previous sections, we choose the grid to be
$202 \times 3001$. Then $5$ lattice Boltzmann units do not occupy more than $60$
percent of the effective film thickness. Because the simulation gets unstable with
smaller grids and larger gradients, all the capillary number simulations were
performed on the same grid. To properly initialize the body force, the
proportionality law was utilized. The forcing
$6 \cdot 10^{-6}/16$ was chosen to obtain the predicted capillary
number $0.05$. 
The pressure gradient can be obtained through the capillary number
ratio:
\begin{equation}
\frac{\mathrm{d}P}{\mathrm{d} x}=6 \cdot 10^{-6}/16 \frac{\Ca_{\mathrm{lit}}}{0.05}
\end{equation}
The film thickness is initialized through the ratio of capillary numbers as well:
\begin{equation*}
w=12 \frac{\Ca_{\mathrm{lit}}}{0.05}
\end{equation*}
The results obtained after $2\cdot10^5$ steps are presented in Table
\ref{table:parameters:capillary:number}. 
{\color{red} We chose to measure the film thickness in the middle of the bubble. The
classical Bretherton formulation is for the measurement of the film thickness at ``infinity''.
However, \citet{cerro-bubble-train} indicate that for the gravity-driven bubble train flow in tubes
it is enough for bubble to have a length of two-three diameters to apply analysis for very long
bubbles. In comparison with the axisymmetric circular geometry, for three-dimensional square shaped
microchannels \citet{heil-threedim} indicate the measurement of the film thickness at $5.5$
channel heights from the front bubble tip. This distance was shown sufficient for the
film thickness to establish itself for the capillary number range $\Ca<4$.
\citet{giavedoni-numerical} showed that the film stabilizes at the distances as $2.6-4.0$ diameters
from the front tip depending on the Reynolds number.  We chose to measure the film thickness in the
middle of the bubble, which is located at least at the distance of $2.5-3$ channel heights from the
front bubble tip. Moreover, the film thickness examination, see Section
\ref{section:film:thickness:variation}, shows that for the capillary range of interest, i.e.
$0.05\leq \Ca \leq 1$, the standard deviation from the bubble middle film thickness is around $7$
percent for small capillary numbers ($\Ca=0.05$) and less than $1$ percent for larger ones ($\Ca
\geq 0.1$) at the distances ranging from one channel height from the rear tip upto one channel
height to the front tip. This certainly validates the point of choice to measure bubble thickness.}
One can see that the calculated
capillary numbers are overpredicted but the obtained thicknesses are overpredicted
as well due to the body force setting caused by the Poiseuille flow assumption. If one
needs to obtain the desired capillary number, the shooting method for the forcing is
necessary, starting with the Poiseuille pressure gradient force as the initial condition. We extracted
the data of 
\citet{giavedoni-numerical} and \citet{heil-bretherton} and compared them with our results (see 
Fig. \ref{fig:capillary:comparison}).
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/Capillary/capillaries_comparison_real.eps}
\caption{A comparison between simulation results and results of
\citet{giavedoni-numerical} and \citet{heil-bretherton}. One can see a
reasonable agreement. The plots depict the film thickness as a function of the
the capillary number.\label{fig:capillary:comparison}}
\end{figure}
